plyranges: a grammar for manipulating genomics data

### 2018-02-13 paper.Rmd
design section mostly complete; now onto results

bibliography: references.bib
output: rticles::plos_article
csl: plos.csl
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
                      fig.width = 6, fig.height = 4)
library(ggbio)
library(ggplot2)
library(plyranges)
```
# Introduction

@@

Several other DSLs have been implemented to reason about genomics data. Broadly, 
these are either implemented as query languages or as command line tools 
embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. An example of the former is the Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
Another  example is the Genometric Query Language (GMQL) which implements a relational algebra for combining big genomic datasets @Kaitoua2017-pw.  The command line application 
BEDtools develops an extensive algebra for performing arithmetic between two or 
more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
which simplifies construcing scripts for performing analyses based on BEDTools  @Dale2011-js. ^[probably should also mention something about their cons, and more detaiil]

@@

In order to provide a compatible API with _dplyr_, _plyranges_ makes use
of non-standard evaluation in R via the _rlang_ package. 
<!--- need to mention something about how this could be detrimental --->

@@

The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
for constructing Ranges from tabular data structures, such as data.frames in
base R. These methods use non-standard evaluation so columns in a data.frame
can be modified before a Ranges object is created. The API also has
convienence methods for annotating or extracting annotations from Genomic Ranges
objects with the `set_genome_info()` and `get_genome_info()` functions.

@@

As an example, we use the Bioconductor package AnnotationHub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project.
We choose to focus on assays for primary T CD8+ memory cells from peripheral 
blood. We can then read the BigWig file  corresponding to the H3K27Me3 
methylation mark over chromosome 10. First we gather the BigWig file
and extract its annotation information and filter it to chromosome 10. 
Then we read the BigWig file only extracting scores if they overlap chromosome
10. We then add annotation information to ensure we know we are working
with the hg19 genome build.

@@

chr10_ranges <- bw_file %>% 
  get_genome_info() %>%
  filter(seqnames == "chr10")

chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
chr10_scores
```

@@

The _plyranges_ API exports the six core verbs from the _dplyr_ package
and modifys them for use with simple _Ranges_ and _Genomic Ranges_. The
verb `mutate()` takes a Ranges object and a set of name-value pairs
and generates a new Ranges object that with modified or new metadata
columns or modified core components (start, end, width, seqnames, strand).
The use of `mutate()` means that a user no longer needs knowledge of the 
accessors of the Ranges object, as they can modify them in place.
The `filter()` function takes a Ranges object and logical expressions and
restricts Ranges object to where the logical expression evaluates to true.
The `summarise()` function takes a Ranges object and a set of name-value pairs
and aggregates the Ranges according to functions evaluated in the name-value pairs.
As `summarise()` is an aggregation it may break the structure of the of a Ranges
object, hence it returns a DataFrame object.
The `select()` function determines
which metadata columns are returned and the order they are returned in. The 
`arrange()` function sorts a Ranges object by named variables.
The `group_by()` function creates an implicit grouping of Ranges object according
to variables in the Ranges object. This modifies the actions of `mutate()`,
`summarise()` and `filter()` (and also the set operations, `reduce_ranges()`
and `disjoin_ranges()`), so they are performed on each partition created by the
grouping.

@@

Returning to the Ranges object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position.

@@

A common operation to perform between two _Ranges_ objects is to find overlaps
or nearest neighbours. The _plyranges_ API recasts these operations as 'joins'
or 'pairing' operations. For overlaps, there are three join operations: 
`join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
which are shown in figure (). These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges objects
as metadata. The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject Ranges 
(for GenomicRanges it also accounts for sequence name), when there is a overlap
the metadata corresponding to the query and subject Ranges are returned. 
Similarly, inner join takes the start and end coordinates of the query Ranges
that overlap the subject Ranges and returns metadata of the overlapping query
and subject Ranges. Finally, the left join performs a left outer join between
the query and subject Ranges, it returns all genomic intervals from the query
ranges, and returns missing values in metadata columns when there is no overlap.

The overlap innner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contains
the max peak on chromosome 10 (visualised in figure ()).

```{r peak_region, fig.align = "center", fig.cap = "Visualisation of peak coverage score accross 5000nt region of chromosome 10"}
peak_region <- chr10_scores %>%
  join_overlap_inner(., 
                     chr10_max_score_region %>% 
                       select(-score))
ggbio::autoplot(peak_region, 
                aes(y = score), 
                geom = "blank") + 
  geom_area(aes(x = start))
```

@@

@@

The _plyranges_ package is avaiable on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the `plyranges` package and extend it for use
with more complex data structures  such as the 
_SummarizedExperiment_ class, which can be used for analysing transcriptomic
and variant data. As the `plyranges` interface encourages tidy data 
practices it integrates well with the principles of the grammar of graphics, 
we aim to use it for the visualisation of genomic ranges.

>>>>

bibliography: references.bib
output: 
  rticles::plos_article:
csl: plos.csl
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, comment = "#>", 
                      fig.width = 5, fig.height = 3, fig.align = "center")
options(digits = 3)
library(ggbio)
library(ggplot2)
library(plyranges)
```
# Introduction

@@

Several other DSLs have been implemented to reason about genomics data. Broadly, 
these are either implemented as query languages or as command line tools 
embedded in the unix environment ^[other ideas to mention GROK @Ovaska2013-gd]. 
An example of the former is the Genome Query Language (GQL) and its distributed 
implementation GenAp which use an SQL-like syntax for fast retrieval of 
information from genomic databases and BAM files @Kozanitis2014-va; @Kozanitis2016-bm. 
Another  example is the Genometric Query Language (GMQL) which implements a 
relational algebra for combining big genomic datasets @Kaitoua2017-pw.  
The command line application 
BEDtools develops an extensive algebra for performing arithmetic between two or 
more sets of genomic regions  @Quinlan2010-gc. It also has a python interface 
which simplifies constructing scripts for performing analyses based on BEDTools 
@Dale2011-js. 
<!--- probably should also mention something about their cons, and more detail --->

@@

In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
use of non-standard evaluation in R via the _rlang_ package. Simply, this
means that compuations are performed and evaluated in the context of the 
Ranges objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with _plyranges_ becomes more difficult 
because a user needs to capture expressions inside function calls and then
unquote them.
<!--- need to make this clearer? --->

@@

The _plyranges_ API provides the methods `as_granges()` and `as_iranges()`
for constructing Ranges from tabular data structures, such as the _data.frame_ in
base R. These methods use non-standard evaluation so columns in a _data.frame_
can be modified before a _Ranges_ object is created. The API also has
convienence methods for annotating or extracting annotations from Genomic Ranges
objects with the `set_genome_info()` and `get_genome_info()` functions.

@@

As an example, we use the Bioconductor package AnnotationHub to search for 
BigWig files from ChIP-Seq experiments from the Human Epigenome Roadmap project.
We choose to focus on assays for primary T CD8+ memory cells from peripheral 
blood. We can then read the BigWig file  corresponding to the H3 lysine 27 
 trimethylation (H3K27Me3) 
methylation mark over chromosome 10. First we gather the BigWig file
and extract its annotation information and filter it to chromosome 10.

@@

chr10_ranges <- bw_file %>% 
  get_genome_info() %>%
  filter(seqnames == "chr10")
```

Then we read the BigWig file only extracting scores if they overlap chromosome
10. We then add annotation information to ensure we know we are working
with the hg19 genome build.

```{r}
chr10_scores <- bw_file %>%
  read_bigwig(overlap_ranges = chr10_ranges) %>%
  set_genome_info(genome = "hg19")
chr10_scores
```

@@

The _plyranges_ API exports the six core verbs from the _dplyr_ package
and modifys them for use with simple _Ranges_ and _Genomic Ranges_. The
verb `mutate()` takes a Ranges object and a set of name-value pairs
and generates a new Ranges object that with modified or new metadata
columns or modified core components (start, end, width, seqnames, strand).
The use of `mutate()` means that a user no longer needs knowledge of the 
accessors of the Ranges object, as they can modify them in place.
The `filter()` function takes a Ranges object and logical expressions and
restricts Ranges object to where the logical expression evaluates to true.
The `summarise()` function takes a Ranges object and a set of name-value pairs
and aggregates the Ranges according to functions evaluated in the name-value pairs.
As `summarise()` is an aggregation it may break the structure of the of a Ranges
object, hence it returns a DataFrame object.
The `select()` function determines
which metadata columns are returned and the order they are returned in. The 
`arrange()` function sorts a Ranges object by named variables. 
The `group_by()` function creates an implicit grouping of Ranges object according
to variables in the Ranges object. This modifies the actions of `mutate()`,
`summarise()` and `filter()`, so they are performed on each partition created 
by the grouping. The `group_by()` operation acts as a replacement for 
the _GenomicRangesList_ class in the original _GenomicRanges_ API.

The _plyranges_ API introduces additional summary verbs, `reduce_ranges()`
and `disjoin_ranges()`that return Ranges objects after being returned. The
`reduce_ranges()` operation merges overlapping and neighbour ranges, while
`disjoin_ranges()` expands ranges by taking the union of end points.

The `reduce_ranges()` operation could be used to find coverage peaks 
accross chromosome 10. We can manually set a threshold to restrict genomic
regions to have a coverage score of greater than 8, and
then merge those regions and compute the maximum coverage over all the regions
that were reduced.

```{r}
all_peaks <- chr10_scores %>% 
  filter(score > 8) %>% 
  reduce_ranges(score = max(score))
```

The combination of these verbs together encapsulate most operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user be exposed to new classes. In those packages to perform
the above operation requires users to understand what an run length encoding
object and views object means and perform additional manipulations to return
the result back to a _Ranges_.

@@

Returning to the Ranges object containing normalised coverage scores for
the methylation data, we can filter to find the coordinates of the peak 
containing maximum coverage score. We can then find 
a 5000 nt region centered around the maximum position by anchoring and 
modifying the the width.

@@

A common operation to perform between two _Ranges_ objects is to find overlaps
or nearest neighbours. The _plyranges_ API recasts these operations as 'joins'
or 'pairing' operations. For overlaps, there are three join operations: 
`join_overlap_intersect()`, `join_overlap_inner()` and `join_overlap_left()`
which are shown in figure (\ref{fig:olap}).

```{r olap, echo = FALSE, fig.cap="The three overlap joins: the query and subject ranges are coloured by their metadata. When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.", fig.width = 5, fig.height = 7}
a <- GRanges(seqnames = "chr1",
               strand = c("+", "-"),
               ranges = IRanges(c(1, 9), c(7, 10)),
               key.a = letters[1:2])

b <- GRanges(seqnames = "chr1",
               strand = c("-", "+"),
               ranges = IRanges(c(2, 6), c(4, 8)),
               key.b = LETTERS[1:2])

c <- join_overlap_intersect(a,b)
d <- join_overlap_inner(a, b)
e <- join_overlap_left(a, b)

p1 <- autoplot(a, aes(fill = key.a)) + 
  scale_fill_brewer(type = "qual", palette = "Set2") + 
  guides(fill = FALSE)
p2 <- autoplot(b, aes(fill = key.b))  + 
  scale_fill_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE)
p3 <- autoplot(c, aes(fill = key.a, colour = key.b)) + 
  scale_fill_brewer(type = "qual", palette = "Set2") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE, color = FALSE) 
p4 <- autoplot(d, aes(fill = key.a, colour = key.b)) + 
  scale_fill_brewer(type = "qual", palette = "Set2") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE, colour = FALSE)
p5 <- autoplot(e, aes(fill = key.a, colour = key.b))  +
  scale_fill_brewer(type = "qual", palette = "Set2") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  guides(fill = FALSE, colour = FALSE)

tracks(query = p1, subject = p2, 
       `intersect` = p3, 
       `inner` = p4, 
       `left` = p5) 
```

These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges objects
as metadata. The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject Ranges 
(for GenomicRanges it also accounts for sequence name), when there is a overlap
the metadata corresponding to the query and subject Ranges are returned. 
Similarly, inner join takes the start and end coordinates of the query Ranges
that overlap the subject Ranges and returns metadata of the overlapping query
and subject Ranges. Finally, the left join performs a left outer join between
the query and subject Ranges, it returns all genomic intervals from the query
ranges, and returns missing values in metadata columns when there is no overlap.

The overlap innner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that contains
the max peak on chromosome 10 (visualised in figure \ref{fig:peak-viz}).

```{r peak_region}
peak_region <- chr10_scores %>%
  join_overlap_inner(., 
                     chr10_max_score_region %>% 
                       select(-score))

```

```{r peak-viz, echo = FALSE, fig.cap = "Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project."}
ggbio::autoplot(peak_region, 
                aes(y = score), 
                geom = "blank") + 
  geom_area(aes(x = start))
```

@@

- Redo an analysis in https://www.nature.com/articles/nature14248

in particular section: Trait-associated variants enrich in tissue-specific marks

@@

The _plyranges_ package is avaiable on the Bioconductor project website 
[https://bioconductor.org](https://bioconductor.org) or can be accessed
via Github [https://github.com/sa-lee/plyranges](https://github.com/sa-lee/plyranges).
We aim to continue developing the `plyranges` package and extend it for use
with more complex data structures  such as the 
_SummarizedExperiment_ class, which can be used for analysing transcriptomic
and variant data. As the `plyranges` interface encourages tidy data 
practices it integrates well with the principles of the grammar of graphics, 
we aim to use it for the visualisation of multimodal biological datasets.

### 2018-02-13 paper.tex
design section mostly complete; now onto results

Several other DSLs have been implemented to reason about genomics data.
Broadly, these are either implemented as query languages or as command
line tools embedded in the unix environment.\footnote{other ideas to
  mention GROK {[}2{]}} An example of the former is the Genome Query
Language (GQL) and its distributed implementation GenAp which use an
SQL-like syntax for fast retrieval of information from genomic databases
and BAM files {[}3{]}; {[}4{]}. Another example is the Genometric Query
Language (GMQL) which implements a relational algebra for combining big
genomic datasets {[}5{]}. The command line application BEDtools develops
an extensive algebra for performing arithmetic between two or more sets
of genomic regions {[}6{]}. It also has a python interface which
simplifies construcing scripts for performing analyses based on BEDTools
{[}7{]}.\footnote{probably should also mention something about their
  cons, and more detaiil}

@@

In order to provide a compatible API with \emph{dplyr}, \emph{plyranges}
makes use of non-standard evaluation in R via the \emph{rlang} package.

@@

The \emph{plyranges} API provides the methods \texttt{as\_granges()} and
\texttt{as\_iranges()} for constructing Ranges from tabular data
structures, such as data.frames in base R. These methods use
non-standard evaluation so columns in a data.frame can be modified
before a Ranges object is created. The API also has convienence methods
for annotating or extracting annotations from Genomic Ranges objects
with the \texttt{set\_genome\_info()} and \texttt{get\_genome\_info()}
functions.

@@

\NormalTok{chr10_ranges <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(seqnames }\OperatorTok{==}\StringTok{ "chr10"}\NormalTok{)}

\NormalTok{chr10_scores <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{read_bigwig}\NormalTok{(}\DataTypeTok{overlap_ranges =}\NormalTok{ chr10_ranges) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
\NormalTok{chr10_scores}
\end{Highlighting}
\end{Shaded}

@@

The \emph{plyranges} API exports the six core verbs from the
\emph{dplyr} package and modifys them for use with simple \emph{Ranges}
and \emph{Genomic Ranges}. The verb \texttt{mutate()} takes a Ranges
object and a set of name-value pairs and generates a new Ranges object
that with modified or new metadata columns or modified core components
(start, end, width, seqnames, strand). The use of \texttt{mutate()}
means that a user no longer needs knowledge of the accessors of the
Ranges object, as they can modify them in place. The \texttt{filter()}
function takes a Ranges object and logical expressions and restricts
Ranges object to where the logical expression evaluates to true. The
\texttt{summarise()} function takes a Ranges object and a set of
name-value pairs and aggregates the Ranges according to functions
evaluated in the name-value pairs. As \texttt{summarise()} is an
aggregation it may break the structure of the of a Ranges object, hence
it returns a DataFrame object. The \texttt{select()} function determines
which metadata columns are returned and the order they are returned in.
The \texttt{arrange()} function sorts a Ranges object by named
variables. The \texttt{group\_by()} function creates an implicit
grouping of Ranges object according to variables in the Ranges object.
This modifies the actions of \texttt{mutate()}, \texttt{summarise()} and
\texttt{filter()} (and also the set operations,
\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}), so they are
performed on each partition created by the grouping.

@@

A common operation to perform between two \emph{Ranges} objects is to
find overlaps or nearest neighbours. The \emph{plyranges} API recasts
these operations as `joins' or `pairing' operations. For overlaps, there
are three join operations: \texttt{join\_overlap\_intersect()},
\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
are shown in figure (). These operations consider any overlap between
two input ranges and return any corresponding metadata from both Ranges
objects as metadata. The intersect join takes the intersect of the start
and end coordinates of overlapping intervals of the query and subject
Ranges (for GenomicRanges it also accounts for sequence name), when
there is a overlap the metadata corresponding to the query and subject
Ranges are returned. Similarly, inner join takes the start and end
coordinates of the query Ranges that overlap the subject Ranges and
returns metadata of the overlapping query and subject Ranges. Finally,
the left join performs a left outer join between the query and subject
Ranges, it returns all genomic intervals from the query ranges, and
returns missing values in metadata columns when there is no overlap.

The overlap innner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that
contains the max peak on chromosome 10 (visualised in figure ()).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(., }
\NormalTok{                     chr10_max_score_region }\OperatorTok{%>%}\StringTok{ }
\StringTok{                       }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{score))}
\NormalTok{ggbio}\OperatorTok{::}\KeywordTok{autoplot}\NormalTok{(peak_region, }
                \KeywordTok{aes}\NormalTok{(}\DataTypeTok{y =}\NormalTok{ score), }
                \DataTypeTok{geom =} \StringTok{"blank"}\NormalTok{) }\OperatorTok{+}\StringTok{ }
\StringTok{  }\KeywordTok{geom_area}\NormalTok{(}\KeywordTok{aes}\NormalTok{(}\DataTypeTok{x =}\NormalTok{ start))}
\end{Highlighting}
\end{Shaded}

@@

{\centering \includegraphics{paper_files/figure-latex/peak_region-1}

@@

\caption{Visualisation of peak coverage score accross 5000nt region of chromosome 10}\label{fig:peak_region}
\end{figure}

@@

@@

The \emph{plyranges} package is avaiable on the Bioconductor project
website \url{https://bioconductor.org} or can be accessed via Github
\url{https://github.com/sa-lee/plyranges}. We aim to continue developing
the \texttt{plyranges} package and extend it for use with more complex
data structures such as the \emph{SummarizedExperiment} class, which can
be used for analysing transcriptomic and variant data. As the
\texttt{plyranges} interface encourages tidy data practices it
integrates well with the principles of the grammar of graphics, we aim
to use it for the visualisation of genomic ranges.

>>>>

Several other DSLs have been implemented to reason about genomics data.
Broadly, these are either implemented as query languages or as command
line tools embedded in the unix environment.\footnote{other ideas to
  mention GROK {[}2{]}} An example of the former is the Genome Query
Language (GQL) and its distributed implementation GenAp which use an
SQL-like syntax for fast retrieval of information from genomic databases
and BAM files {[}3{]}; {[}4{]}. Another example is the Genometric Query
Language (GMQL) which implements a relational algebra for combining big
genomic datasets {[}5{]}.\\
The command line application BEDtools develops an extensive algebra for
performing arithmetic between two or more sets of genomic regions
{[}6{]}. It also has a python interface which simplifies constructing
scripts for performing analyses based on BEDTools {[}7{]}.

@@

In order to provide a compatible API with \emph{dplyr}, \emph{plyranges}
makes extensive use of non-standard evaluation in R via the \emph{rlang}
package. Simply, this means that compuations are performed and evaluated
in the context of the Ranges objects, which emphasises the interactive
nature of the API. This has the disadvantage that programming with
\emph{plyranges} becomes more difficult because a user needs to capture
expressions inside function calls and then unquote them.

@@

The \emph{plyranges} API provides the methods \texttt{as\_granges()} and
\texttt{as\_iranges()} for constructing Ranges from tabular data
structures, such as the \emph{data.frame} in base R. These methods use
non-standard evaluation so columns in a \emph{data.frame} can be
modified before a \emph{Ranges} object is created. The API also has
convienence methods for annotating or extracting annotations from
Genomic Ranges objects with the \texttt{set\_genome\_info()} and
\texttt{get\_genome\_info()} functions.

@@

\NormalTok{chr10_ranges <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{get_genome_info}\NormalTok{() }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(seqnames }\OperatorTok{==}\StringTok{ "chr10"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

Then we read the BigWig file only extracting scores if they overlap
chromosome 10. We then add annotation information to ensure we know we
are working with the hg19 genome build.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{chr10_scores <-}\StringTok{ }\NormalTok{bw_file }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{read_bigwig}\NormalTok{(}\DataTypeTok{overlap_ranges =}\NormalTok{ chr10_ranges) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{set_genome_info}\NormalTok{(}\DataTypeTok{genome =} \StringTok{"hg19"}\NormalTok{)}
\NormalTok{chr10_scores}
\end{Highlighting}
\end{Shaded}

@@

The \emph{plyranges} API exports the six core verbs from the
\emph{dplyr} package and modifys them for use with simple \emph{Ranges}
and \emph{Genomic Ranges}. The verb \texttt{mutate()} takes a Ranges
object and a set of name-value pairs and generates a new Ranges object
that with modified or new metadata columns or modified core components
(start, end, width, seqnames, strand). The use of \texttt{mutate()}
means that a user no longer needs knowledge of the accessors of the
Ranges object, as they can modify them in place. The \texttt{filter()}
function takes a Ranges object and logical expressions and restricts
Ranges object to where the logical expression evaluates to true. The
\texttt{summarise()} function takes a Ranges object and a set of
name-value pairs and aggregates the Ranges according to functions
evaluated in the name-value pairs. As \texttt{summarise()} is an
aggregation it may break the structure of the of a Ranges object, hence
it returns a DataFrame object. The \texttt{select()} function determines
which metadata columns are returned and the order they are returned in.
The \texttt{arrange()} function sorts a Ranges object by named
variables. The \texttt{group\_by()} function creates an implicit
grouping of Ranges object according to variables in the Ranges object.
This modifies the actions of \texttt{mutate()}, \texttt{summarise()} and
\texttt{filter()}, so they are performed on each partition created by
the grouping. The \texttt{group\_by()} operation acts as a replacement
for the \emph{GenomicRangesList} class in the original
\emph{GenomicRanges} API.

The \emph{plyranges} API introduces additional summary verbs,
\texttt{reduce\_ranges()} and \texttt{disjoin\_ranges()}that return
Ranges objects after being returned. The \texttt{reduce\_ranges()}
operation merges overlapping and neighbour ranges, while
\texttt{disjoin\_ranges()} expands ranges by taking the union of end
points.

The \texttt{reduce\_ranges()} operation could be used to find coverage
peaks accross chromosome 10. We can manually set a threshold to restrict
genomic regions to have a coverage score of greater than 8, and then
merge those regions and compute the maximum coverage over all the
regions that were reduced.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{all_peaks <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{filter}\NormalTok{(score }\OperatorTok{>}\StringTok{ }\DecValTok{8}\NormalTok{) }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{reduce_ranges}\NormalTok{(}\DataTypeTok{score =} \KeywordTok{max}\NormalTok{(score))}
\end{Highlighting}
\end{Shaded}

The combination of these verbs together encapsulate most operations that
can be performed in the original \emph{IRanges} and \emph{GenomicRanges}
packages without the user be exposed to new classes. In those packages
to perform the above operation requires users to understand what an run
length encoding object and views object means and perform additional
manipulations to return the result back to a \emph{Ranges}.

@@

A common operation to perform between two \emph{Ranges} objects is to
find overlaps or nearest neighbours. The \emph{plyranges} API recasts
these operations as `joins' or `pairing' operations. For overlaps, there
are three join operations: \texttt{join\_overlap\_intersect()},
\texttt{join\_overlap\_inner()} and \texttt{join\_overlap\_left()} which
are shown in figure (\ref{fig:olap}).

@@

{\centering \includegraphics{paper_files/figure-latex/olap-1}

@@

\caption{The three overlap joins: the query and subject ranges are coloured by their metadata. When an overlap is performed the resulting range is filled by the query metadata and the metadata from the subject colours the outside of the range.}\label{fig:olap}
\end{figure}

These operations consider any overlap between two input ranges and
return any corresponding metadata from both Ranges objects as metadata.
The intersect join takes the intersect of the start and end coordinates
of overlapping intervals of the query and subject Ranges (for
GenomicRanges it also accounts for sequence name), when there is a
overlap the metadata corresponding to the query and subject Ranges are
returned. Similarly, inner join takes the start and end coordinates of
the query Ranges that overlap the subject Ranges and returns metadata of
the overlapping query and subject Ranges. Finally, the left join
performs a left outer join between the query and subject Ranges, it
returns all genomic intervals from the query ranges, and returns missing
values in metadata columns when there is no overlap.

The overlap innner join could be used to restrict the chromosome 10
normalised coverage scores that are within the 5000nt region that
contains the max peak on chromosome 10 (visualised in figure
\ref{fig:peak-viz}).

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{peak_region <-}\StringTok{ }\NormalTok{chr10_scores }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{join_overlap_inner}\NormalTok{(., }
\NormalTok{                     chr10_max_score_region }\OperatorTok{%>%}\StringTok{ }
\StringTok{                       }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{score))}
\end{Highlighting}
\end{Shaded}

{\centering \includegraphics{paper_files/figure-latex/peak-viz-1}

\caption{Visualisation of normalised coverage scores accross a 5000nt region of chromosome 10 from H3K27Me3 ChIP-Seq assay from the Human Epigenome Roadmap project.}\label{fig:peak-viz}
\end{figure}

@@

\begin{itemize}
\tightlist
\item
  Redo an analysis in https://www.nature.com/articles/nature14248
\end{itemize}

@@

The \emph{plyranges} package is avaiable on the Bioconductor project
website \url{https://bioconductor.org} or can be accessed via Github
\url{https://github.com/sa-lee/plyranges}. We aim to continue developing
the \texttt{plyranges} package and extend it for use with more complex
data structures such as the \emph{SummarizedExperiment} class, which can
be used for analysing transcriptomic and variant data. As the
\texttt{plyranges} interface encourages tidy data practices it
integrates well with the principles of the grammar of graphics, we aim
to use it for the visualisation of multimodal biological datasets.
